Leadership across cultures and contexts: an exploratory analysis

### 2020-04-07 Response to reviewers.Rmd

We added a few additional details in the methods section on the eHRAF and ethnographic sample and now more clearly direct readers to the original publication from which the database was built, which includes additional details.

We also include additional details in the results and supplementary information on the corpus of ethnographic documents which provide the sample of ethnographic paragraphs. NEED TO DO.

**I don't work with ethnographic data, so perhaps this question is naïve, but: What conclusions can we draw about the actual frequency of leadership functions and traits from the frequency with which they are mentioned in the ethnographic record? There is an implicit assumption that the two are closely aligned. Has such an alignment been confirmed? Is that even testable?**

The reviewer brings up an important point. For one, this is partly our motivation for representing the percentage of cultures which provide at least one text record supporting each of the operationalized variables. Despite this low threshold for accepting supporting evidence, the ethnographic record (sampled here) does not provide support for every variable from every culture. At this culture-level we feel more confident our data provide a estimate of cross-cultural frequency, albeit an inherently noisy, "course-grained" estimation. Also, the entire field of ethnology (the branch of anthropology focused on comparing characteristics of populations across the globe) is based on developing comparable and often quantitative measures from primary ethnographic information. For example, the Ethnographic Atlas, which provides hundreds of coded quantitative and qualitative variables on thousands of cultures and the Standard Cross-Cultural Sample, which provides thousands of coded quantitative and qualitative variables on 186 cultures, were developed in essentially the same way we have developed our operationalized variables here. Developing and contrasting coded measures from primary ethnographic documents is not without limitations, but does represent (perhaps) the only method to systematically and empirically compare measures across a large sample of cultures/populations. At the text-record level drawing strong conclusions about the actual frequency of traits is slightly more problematic. However, fundamentally these results provide a systematic view of the ethnographic record on these measures. Our intention is that from our results field researchers and theoreticians interested in these topics can draw inference from a more comprehensive view of human cultural diversity rather than haphazardly self-selecting and relying on only a handful of ethnographic cases studies.

@@

We greatly appreciate the detailed comments and very useful suggestions.

**The abstract could be tightened up in terms of description of particular costs and benefits to leaders and followers. And a brief justification in the abstract for why the study compares costs and benefits would be useful- e.g. understanding evolution of leader or follower strategies requires understanding the (fitness-relevant) costs and benefits to them.**

**Lines 88-89 suggest that the paper's ethnographic analysis will include leadership at all levels of social organization. But does the theory and analysis reported in the paper really extend to all levels of social organization- such as child play groups or households? It seems the focus is still community-level cooperation among adults. Perhaps clarify.**

This statement in the introduction is intended to apply across disciplines focused on leadership dynamics and not specifically apply to the methods of the current study. The reviewer brings up a well-recieved point, however. Any work in developmental psychology on social hierarchy adn leadership would not be subjec to this critique. We have not rephrased this claim to better fit with our objectives and outcomes here.

@@

**While your analysis strategy was not designed to test particular predictions, I'd be very interested to see you report how the authority variable covaries with different leader functions in different subsistence types, which has relevance to institutionalization of leadership and evolution of social inequality.**

**Give more explanation of prestige theory and computational services theory (lines 1279-1280), including what different predictions they make.**

**Followers in many cases may be making best of limited options, particularly in societies with steeper hierarchy. So I'm not sure benefits of followership must outweigh the costs (1414-1415) in same way as for leadership.**

We completely agree. This statement was intneded to be in reference to a priori assumptions on the evolution and emergence of leader-follower dynamics, similar to discourse on the conundrum of the evolution of cooperation. Though, clearly, in particular circumstances an individual will cooperate when doing so incurs a cost just as followers in particular circumstances will follow when doing so incurs a cost. We assume that over evolutionary history and for leader-follower dynamics to be selected for, both leaders and followers would have to recieve benefits on average, over time.

We have revised this statement. CHECK. NEED TO DO.

### 2020-04-07 leadership_across_cultures_contexts.Rmd
Working on reviewer comments and MS.
Added more stop words in text analyses.

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and at all levels of social organization. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have been relatively under-investigated. Current outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is shaped by population-level variation at different levels of social organizations and the costs and benefits leaders and followers incur across distinct leadership systems.

@@

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. In the absence of institutionalized organizational structure leaders often take on essential roles in group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who do not appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. @glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military fromidability to rival groups and facilitating within-group conflict resolution and cooperation. In addition to military prowess, 'big-men' are characterized by, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009][p.104].

@@

There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 found in a systematic cross-cultural sample that in only seven of 93 societies political leaders of both sexes were present and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group. Women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. More recently, empirical studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. Ethnographic fieldwork comparing leaders and non-leaders in two distinct small-scale societies has revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

@@

@@

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership must also outweigh the costs. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits and social status, but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

>>>>

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have been relatively under-investigated. Current outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation at different levels of social organizations and the costs and benefits leaders and followers incur across distinct leadership systems.

@@

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. In the absence of institutionalized organizational structure leaders often take on essential roles in group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who do not appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. @glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military fromidability to rival groups and facilitating within-group conflict resolution and cooperation. In addition to military prowess, 'big-men' are characterized by, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

@@

There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary discussion and consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 and coneptualize social status as a relative indicator of an individuals social value by the group and their access to contested resources within the group. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [@von_rueden_leadership_2015; @van_vugt_genes_2020], however, if there is causality between the two it is likely to be complex, context-specific, and driven by underlying correlates.

Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 found in a systematic cross-cultural sample that in only seven of 93 societies political leaders of both sexes were present and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group. Women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. More recently, empirical studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. Ethnographic fieldwork comparing leaders and non-leaders in two distinct small-scale societies has revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

@@

To suppliment results from the aformentioend methods, we also performed textual analyses...OTHER METHODS. NEED TO DO.

@@

## Disentangleing the relationship between leadership and high status

Figure \@ref(fig:varsupport) identified *high status* as the most frequent leader quality in the ethnographic record, documented in over 30% of text records and in over 90% of cultures. This result is neither surprising nor particularly informative. However, these data present an oppurtunity to explore the relationship between high status and leadership, as represented in ethnographic texts. To do so, we draw on the term-document matrix of our ethnographic sample which was developed by computing the frequencies of all “informative” words in each text record after removing all punctuation and English stop words (uninformative words such as the, is, at), stemming the remaining words (reducing inflections such as plurals and past tenses to the root, or stem, words). In this data matrix the columns are all the unique word stems (terms), rows were the text records (documents), and the values in each cell were the frequencies of each word stem in each text record. The coding of *high status* was joined with this term-document matrix. This process produced a matrix with `r dim(dtm)[1]` rows with one for each text-record and `r dim(dtm)[2]` columns with a column for each unique term and a column for the researcher-coded *high status* value ($0$ or $1$).

We used lasso regression as a variable selection technique, which is especially useful when the number of predictors is much greater than the number of cases (i.e., $p > > n$). Lasso regression has a penalty or shrinkage term, $\lambda$: larger values of $\lambda$ retain more variables in the model, at the risk of over-fitting, whereas smaller values of $\lambda$ decrease the number of predictors retained in the model, at the risk of under-fitting. The optimal value of $\lambda$ can be estimated with cross-validation [@tibshirani_regression_1996; @zou_regularization_2005]. We used the glmnet package [@friedman_regularization_2010] to fit a lasso regression model to determine which words best predicted evidence for high status as a leader quality. Because this outomce is counts of text recrods, we used Poisson errors. We used tenfold cross-validation to choose the $\lambda$ parameter that minimized deviance. We then plotted all the non-zero coefficients (Figure \@ref(fig:status-text). The most predictive words then epitomize the semantic content of the text records which provide evidence of high status as a quality of leaders.

```{r status-text, echo=F, message=F, fig.width=8, fig.height=6, fig.align='center', fig.cap="Lasso Poisson regression coefficients, which indicate the words whose frequencies in each text record best predicted evidence for the *high status* leader quality in each text record. Coefficients estimated with glmnet package using cross-validation to determine the optimal value of lambda [@friedman_regularization_2010]. Blue values are positive predictors; red values are negative predictors."}
highstatus_plot
```

Figure \@ref(fig:status-text) depicts the `r length(highstatus_plot$data$Word)` words with non-zero (mostly positive) coefficents. The words most predictive of evidence for high status leaders are *`r rev(highstatus_plot$data$Word)[1:3]`*, and *`r rev(highstatus_plot$data$Word)[4]`*, respectively. The words, *`r rev(highstatus_plot$data$Word)[5:7]`* are moderately strong predictors and notable words with low predictive value (including positive and negative predictors) include *authority*, *influence*, *king*, *chief*, and *headman*. The word *follow* is the strongest negative predictor and *company* and *economic* are marginal negative predictors.

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership (and by extension, group living) must also outweigh the costs for leadership to evolve and be maintained. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits and social status, but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).
